---
title: "Hands-on Exercise 1: Geospatial Data Wrangling with R"
editor: visual
---

## Overview

In this hands on exercise, I learnt the aspects of importing and wrangling geospatial data using appropriate R packages.

## Getting Started

The code chunk below installs and loads sf and tidyverse packages into R environment.

```{r}
pacman::p_load(sf,tidyverse)
```

## Import Geospatial Data

### Import polygon feature data

```{r}
mpsz <- st_read(dsn="data/geospatial",layer = "MP14_SUBZONE_WEB_PL")
```

### Import polyline feature data

```{r}
cyclingpath = st_read(dsn = "data/geospatial", 
                         layer = "CyclingPath")
```

Import GIS data in kml format

```{r}
preschool = st_read("data/geospatial/pre-schools-location-kml.kml")
```

## Check content of simple feature data frame

### Working with st_geometry

```{r}
st_geometry(mpsz)
```

### Working with glimpse

```{r}
glimpse(mpsz)
```

### Working with head()

```{r}
head(mpsz, n=5)  
```

## Plotting the geospatial data

```{r}
plot(mpsz)
```

```{r}
plot(st_geometry(mpsz))
```

```{r}
plot(mpsz["PLN_AREA_N"])
```

## Working with projection

### Assign EPSG code to a simple feature data frame

```{r}
st_crs(mpsz)
```

```{r}
mpsz3414 <- st_transform(mpsz, 3414)
```

```{r}
st_crs(mpsz3414)
```

### Transforming the projection of preschool from wgs84 to svy21

```{r}
preschool3414 <- st_transform(preschool, 
                              crs = 3414)
```

```{r}
st_geometry(preschool3414)
```

## Importing and converting aspatial data

### Importing aspatial data

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

```{r}
list(listings) 
```

### Creating a simple feature data frame from an aspatial data frame

```{r}
listings_sf <- st_as_sf(listings, 
                       coords = c("longitude", "latitude"),
                       crs=4326) %>%
  st_transform(crs = 3414)
```

```{r}
glimpse(listings_sf)
```

## Geoprocessing with sf package

### buffering

```{r}
buffer_cycling <- st_buffer(cyclingpath, 
                               dist=5, nQuadSegs = 30)
```

```{r}
buffer_cycling$AREA <- st_area(buffer_cycling)
```

```{r}
sum(buffer_cycling$AREA)
```

### Point-in-polygon count

```{r}
mpsz3414$`PreSch Count`<- lengths(st_intersects(mpsz3414, preschool3414))

```

```{r}
summary(mpsz3414$`PreSch Count`)
```

```{r}
top_n(mpsz3414, 1, `PreSch Count`)
```

```{r}
mpsz3414$Area <- mpsz3414 %>%
  st_area()
```

```{r}
mpsz3414 <- mpsz3414 %>%
  mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)
```

Exploratory Data Analysis

```{r}
hist(mpsz3414$`PreSch Density`)
```

```{r}
ggplot(data=mpsz3414, 
       aes(x= as.numeric(`PreSch Density`)))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  labs(title = "Are pre-school even distributed in Singapore?",
       subtitle= "There are many planning sub-zones with a single pre-school, on the other hand, \nthere are two planning sub-zones with at least 20 pre-schools",
      x = "Pre-school density (per km sq)",
      y = "Frequency")
```

```{r}
library(ggplot2)
ggplot(data=mpsz3414, aes(x= as.numeric(`PreSch Density`),y=as.numeric(`PreSch Count`))) + geom_point() + labs(x = "Pre-school density (per km sq)",y = "Pre-school count")
```
