---
title: "Hands-on Exercise 1: Geospatial Data Wrangling with R"
editor: visual
---

## Overview

In this hands on exercise, I learnt the aspects of importing and wrangling geospatial data using appropriate R packages.

## Getting Started

The code chunk below installs and loads sf and tidyverse packages into R environment.

```{r}
pacman::p_load(sf,tidyverse)
```

## Import Geospatial Data

### Import polygon feature data

```{r}
mpsz <- st_read(dsn="data/geospatial",layer = "MP14_SUBZONE_WEB_PL")
```

### Import polyline feature data

```{r}
cyclingpath = st_read(dsn = "data/geospatial", 
                         layer = "CyclingPath")
```

Import GIS data in kml format

```{r}
preschool = st_read("data/geospatial/pre-schools-location-kml.kml")
```

## Check content of simple feature data frame

### Working with st_geometry

```{r}
st_geometry(mpsz)
```

### Working with glimpse

```{r}
glimpse(mpsz)
```

### Working with head()

```{r}
head(mpsz, n=5)  
```

## Plotting the geospatial data

```{r}
plot(mpsz)
```

```{r}
plot(st_geometry(mpsz))
```

```{r}
plot(mpsz["PLN_AREA_N"])
```

## Working with projection

### Assign EPSG code to a simple feature data frame

```{r}
st_crs(mpsz)
```

```{r}
mpsz3414 <- st_transform(mpsz, 3414)
```

```{r}
st_crs(mpsz3414)
```

### Transforming the projection of preschool from wgs84 to svy21

```{r}
preschool3414 <- st_transform(preschool, 
                              crs = 3414)
```

```{r}
st_geometry(preschool3414)
```
