---
title: "In Class Exercise 1"
editor: visual
---

## Overview

### Getting Started

```{r}
packages = c('sf', 'spdep', 'tmap', 'tidyverse')
for (p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p,character.only = T)
}
```

The code chunk below will install and load tidyverse and sf packages.

```{r}
pacman::p_load(sf,tidyverse,spdep,tmap)
```

## Getting Data into R environment

### Importing shapefile into R environment

```{r}
hunan <- st_read(dsn = "data/geospatial/", 
                layer = "Hunan")
```

### Import CSV file into R Environment

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### performing relational join

```{r}
hunan <- left_join(hunan,hunan2012)
```

## Visualising Regional Development Indicator

```{r}
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan, "GDPPC")
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

## Computing Contiguity Spatial Weights

### Computing (queen) contiguity based neighbours

```{r}
wm_q <- poly2nb(hunan, queen=TRUE)
summary(wm_q)
```

```{r}
wm_q[[1]]
```

```{r}
hunan$County[1]
```

```{r}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

```{r}
str(wm_q)
```

### Creating (ROOK) contiguity based neighbours

```{r}
wm_r <- poly2nb(hunan, queen=FALSE)
summary(wm_r)
```

### Visualising contiguity weights

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
```

```{r}
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```

```{r}
coords <- cbind(longitude, latitude)
```

```{r}
head(coords)
```

#### Plotting Queen contiguity based neighbours map

```{r}
plot(hunan$geometry, border="lightgrey")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red")
```

#### Plotting Rook contiguity based neighbours map

```{r}
plot(hunan$geometry, border="lightgrey")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

#### Plotting both Queen and Rook contiguity based neighbours maps

```{r}
par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red", main="Queen Contiguity")
plot(hunan$geometry, border="lightgrey")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red", main="Rook Contiguity")
```

## Computing distance based neighbours

### Determine the cut-off distance

```{r}
#coords <- coordinates(hunan)
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

### Computing fixed distance weight matrix

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

```{r}
str(wm_d62)
```

```{r}
table(hunan$County, card(wm_d62))
```

```{r}
n_comp <- n.comp.nb(wm_d62)
n_comp$nc
```

```{r}
table(n_comp$comp.id)
```

#### Plotting fixed distance based weight matrix

```{r}
plot(hunan$geometry, border="lightgrey")
plot(wm_d62, coords, add=TRUE)
plot(k1, coords, add=TRUE, col="red", length=0.08)
```

```{r}
par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey")
plot(k1, coords, add=TRUE, col="red", length=0.08, main="1st nearest neighbours")
plot(hunan$geometry, border="lightgrey")
plot(wm_d62, coords, add=TRUE, pch = 19, cex = 0.6, main="Distance link")
```

### Computing adaptive distance weight matrix

```{r}
knn6 <- knn2nb(knearneigh(coords, k=6))
knn6
```

```{r}
str(knn6)
```

#### Plotting distance based neighbours

```{r}
plot(hunan$geometry, border="lightgrey")
plot(knn6, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

## Weights based on IDW

```{r}
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids <- lapply(dist, function(x) 1/(x))
ids
```

### Row-standardised weights matrix

```{r}
rswm_q <- nb2listw(wm_q, style="W", zero.policy = TRUE)
rswm_q
```

```{r}
rswm_q$weights[10]
```

```{r}
rswm_ids <- nb2listw(wm_q, glist=ids, style="B", zero.policy=TRUE)
rswm_ids
```

```{r}
rswm_ids$weights[1]
```

```{r}
summary(unlist(rswm_ids$weights))
```

## Application of Spatial Weight Matrix

### Spatial lag with row-standardized weights

```{r}
GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
GDPPC.lag
```

```{r}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

```{r}
lag.list <- list(hunan$NAME_3, lag.listw(rswm_q, hunan$GDPPC))
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan,lag.res)
```

```{r}
head(hunan)
```

```{r}
gdppc <- qtm(hunan, "GDPPC")
lag_gdppc <- qtm(hunan, "lag GDPPC")
tmap_arrange(gdppc, lag_gdppc, asp=1, ncol=2)
```

### Spatial lag as a sum of neighboring values

```{r}
b_weights <- lapply(wm_q, function(x) 0*x + 1)
b_weights2 <- nb2listw(wm_q, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

```{r}
lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")
```

```{r}
lag_sum
```

```{r}
hunan <- left_join(hunan, lag.res)
```

```{r}
gdppc <- qtm(hunan, "GDPPC")
lag_sum_gdppc <- qtm(hunan, "lag_sum GDPPC")
tmap_arrange(gdppc, lag_sum_gdppc, asp=1, ncol=2)
```

### Spatial window average

```{r}
wm_q1 <- wm_q
```

```{r}
include.self(wm_q1)
```

```{r}
wm_q1 <- nb2listw(wm_q1)
wm_q1
```

```{r}
lag_w_avg_gpdpc <- lag.listw(wm_q1, 
                             hunan$GDPPC)
lag_w_avg_gpdpc
```

```{r}
lag.list.knn6a <- list(hunan$NAME_3, lag.listw(wm_q1, hunan$GDPPC))
lag_knn6a.res <- as.data.frame(lag.list.knn6a)
colnames(lag_knn6a.res) <- c("NAME_3", "lag_window_avg GDPPC")
```

### Spatial window sum

```{r}
wm_q1 <- wm_q
```

```{r}
include.self(wm_q1)
```

```{r}
wm_q1
```

```{r}
b_weights <- lapply(wm_q1, function(x) 0*x + 1)
b_weights[1]
```

```{r}
b_weights2 <- nb2listw(wm_q1, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

```{r}
w_sum_gdppc <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
w_sum_gdppc
```

```{r}
w_sum_gdppc.res <- as.data.frame(w_sum_gdppc)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")
```

```{r}
hunan <- left_join(hunan, w_sum_gdppc.res)
```

```{r}
gdppc <- qtm(hunan, "GDPPC")
w_sum_gdppc <- qtm(hunan, "w_sum GDPPC")
tmap_arrange(gdppc, w_sum_gdppc, asp=1, ncol=2)
```
